INFOMDA2: Group Assignment

### 2025-01-16 code/analysis.qmd
final changes to code, add plots

---
title: "INFOMDA2: Group Assignment"
author: "Martijn van Dam, Jonathan Koop, Merlin Urbanski"
format: html
---

@@

```{r}
library(pacman)

p_load("readr", "tidyverse", "caret", "kableExtra")
```

@@

```{r}
data <- read.csv("../raw_data/data.csv")
```

@@

```{r}
head(data)
```

@@

```{r}
str(data)
```

@@

```{r}
data <- data %>%
  mutate(class = as.factor(class)) # change class to factor

data$class <- as.factor(data$class)
```

@@

```{r}
sig_05 <- data %>%
  select(-ID) %>% # remove ID column
  summarise(across(-class, ~ t.test(.x ~ class)$p.value)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  filter(V1 < 0.05)
```

@@

```{r}
sig_01 <- data %>%
  select(-ID) %>% # remove ID column
  summarise(across(-class, ~ t.test(.x ~ class)$p.value)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  filter(V1 < 0.01)
```

@@

```{r}
sig_001 <- data %>%
  select(-ID) %>% # remove ID column
  summarise(across(-class, ~ t.test(.x ~ class)$p.value)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  filter(V1 < 0.001)
```

@@

```{r}
set.seed(123)
model_benchmark <- train(class ~ ., data = data_benchmark, method = "glm", family = "binomial", trControl = trainControl(method = "cv", number = 10))

summary(model_benchmark)
model_benchmark$results
```

Even though we have reduced the number of variables to `r nrow(sig_001)`, the algorithm still does not seem to converge. In the li

## 3.3 7 Most Significant Variables

```{r}
sig_7 <- sig_001 %>%
  top_n(7, V1) %>%
  pull(variable)

data_7 <- data %>%
  select(sig_7, class) %>% 
  train(class ~ ., data = ., method = "glm", family = "binomial", trControl = trainControl(method = "cv", number = 10))

data_7$results
```

@@

```{r}
pca <- data %>%
  select(-ID, -class) %>%
  prcomp(center = TRUE, scale. = TRUE)
```

@@

```{r}
screeplot(pca, type = "lines", main = "Scree Plot")
```

@@

```{r}
pca_table %>%
  filter(eigenvalue > 1) %>%
  kable("html") %>%
  kable_styling("striped", full_width = F)
```

@@

```{r}
pc_greaterthan1 <- pca_table %>%
  filter(eigenvalue > 1) %>%
  pull(component) %>% 
  paste0("pc", .)
```

@@

```{r}
data_pca <- bind_cols(
  class = data$class,  # Add class column
  as_tibble(pca$x)     # Convert PCA matrix to tibble
) %>%
  set_names(c("class", paste0("pc", 1:ncol(pca$x)))) 
```

### 4.2.2 Scree Plot

```{r}
logit_scree <- data_pca %>%
  select(pc1, class) %>%
  train(
    class ~ .,
    data = .,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "cv", number = 10)
  )

summary(logit_scree)
logit_scree$results
```

### 4.2.3 Eigenvalue Greater Than 1 Rule

```{r}
logit_greaterthan1 <- data_pca %>% 
  select(all_of(pc_greaterthan1), class) %>%
  train(
    class ~ .,
    data = .,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "cv", number = 10)
  )

summary(logit_greaterthan1)
logit_greaterthan1$results
```

```{r}
logit_7 <- data_pca %>% 
  select(pc1, pc2, pc3, pc4, pc5, pc6, pc7, class) %>%
  train(
    class ~ .,
    data = .,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "cv", number = 10)
  )

summary(logit_7)
logit_7$results
```

>>>>

---
title: "INFOMDA2: Group Assignment"
author: "Martijn van Dam, Jonathan Koop, Merlin Urbanski"
format: html
knitr:
  opts_chunk: 
    warning: false
---

@@

```{r}
# if required, install pacman package
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

library(pacman)

# load all packages with pacman
p_load("readr", "tidyverse", "caret", "kableExtra", "pROC")
```

@@

```{r}
data <- read.csv("../raw_data/data.csv") # load data
```

@@

```{r}
head(data) # show first few rows of data
```

@@

```{r}
str(data) # show structure of data
```

@@

```{r}
# Convert class to factor
data$class <- as.factor(data$class)
```

@@

```{r}
sig_05 <- data %>%
  select(-ID) %>% # remove ID column
  summarise(across(-class, ~ t.test(.x ~ class)$p.value)) %>% # perform t-tests
  t() %>% # transpose
  as.data.frame() %>%
  rownames_to_column("variable") %>% # add variable names as a column
  filter(V1 < 0.05) # filter out variables with p-value < 0.05
```

@@

```{r}
sig_01 <- data %>%
  select(-ID) %>% # remove ID column
  summarise(across(-class, ~ t.test(.x ~ class)$p.value)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  filter(V1 < 0.01) # filter out variables with p-value < 0.01
```

@@

```{r}
sig_001 <- data %>%
  select(-ID) %>% # remove ID column
  summarise(across(-class, ~ t.test(.x ~ class)$p.value)) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("variable") %>%
  filter(V1 < 0.001) # filter out variables with p-value < 0.001
```

@@

#### 3.2.2.1 Training the Model

```{r}
# | warning: false

set.seed(123) # set seed for reproducibility

# Train logistic regression model
model_benchmark <- train(
  class ~ ., # all significant variables
  data = data_benchmark,
  method = "glm",
  family = "binomial",
  trControl = trainControl(method = "LOOCV", savePredictions = "final", classProbs = TRUE)) # leave-one-out cross-validation

summary(model_benchmark) # show summary of model
```

Even though we have reduced the number of variables to `r nrow(sig_001)`, the algorithm still does not seem to converge.

#### 3.2.2.2 Making Predictions and Evaluating the Model

```{r}
# Accuracy and Kappa
model_benchmark$results

# ROC Curve and AUC
predictions_model_benchmark <- model_benchmark$pred
roc_benchmark <- roc(response = predictions_model_benchmark$obs, predictor = predictions_model_benchmark$P)
```

@@

```{r}
pca <- data %>%
  select(-ID, -class) %>% # remove ID and class columns
  prcomp(center = TRUE, scale. = TRUE) # perform PCA
```

@@

```{r}
screeplot(pca, type = "lines", main = "Scree Plot") # plot scree plot
```

@@

```{r}
pca_table %>%
  filter(eigenvalue > 1) %>% # filter out components with eigenvalue > 1
  kable("html") %>% # create HTML table
  kable_styling("striped", full_width = F) # add styling
```

@@

```{r}
pc_greaterthan1 <- pca_table %>% 
  filter(eigenvalue > 1) %>% # filter out components with eigenvalue > 1
  pull(component) %>% # extract component numbers
  paste0("pc", .) # create column names
```

@@

```{r}
data_pca <- bind_cols(
  class = data$class,  # Add class column
  as_tibble(pca$x)     # Convert PCA matrix to tibble
) %>%
  set_names(c("class", paste0("pc", 1:ncol(pca$x)))) # Rename columns
```

### 4.2.2 Logistic Regression Using Elbow Criterion

#### 4.2.2.1 Training the Model

```{r}
model_scree <- data_pca %>%
  select(pc1, class) %>%
  train(
    class ~ .,
    data = .,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "LOOCV", savePredictions = "final", classProbs = TRUE) # leave-one-out cross-validation
  )

summary(model_scree) # show summary of model
```

#### 4.2.2.2 Making Predictions and Evaluating the Model

```{r}
# Accuracy and Kappa
model_scree$results

# ROC Curve and AUC
predictions_model_scree <- model_scree$pred
roc_scree<- roc(response = predictions_model_scree$obs, predictor = predictions_model_scree$P)
```

### 4.2.3 Logistic Regression Using Eigenvalue Greater Than 1 Rule

#### 4.2.3.1 Training the Model

```{r}
model_greaterthan1 <- data_pca %>% 
  select(all_of(pc_greaterthan1), class) %>%
  train(
    class ~ .,
    data = .,
    method = "glm",
    family = "binomial",
    trControl = trainControl(method = "LOOCV", savePredictions = "final", classProbs = TRUE)
  )

summary(model_greaterthan1)
```

#### 4.2.3.2 Making Predictions and Evaluating the Model

```{r}
# Accuracy and Kappa
model_greaterthan1$results

# ROC Curve and AUC
predictions_model_greaterthan1 <- model_greaterthan1$pred
roc_greaterthan1 <- roc(response = predictions_model_greaterthan1$obs, predictor = predictions_model_greaterthan1$P)
```

# 5. Plots

## 5.1 Scree Plot

```{r}
# Slice out the first 90 components
pca_table_90 <- pca_table %>% 
  slice(1:90)

ggplot(pca_table_90, aes(x = component)) +
  # Bar plot for variance explained
  geom_col(aes(y = eigenvalue), fill = "steelblue") +

  # Line + points for cumulative explained variance, scaled to max(eigenvalue)
  geom_line(aes(y = cumulative * max(eigenvalue)), color = "red", size = 1) +
  geom_point(aes(y = cumulative * max(eigenvalue)), color = "red", size = 2) +

  # Set up x-axis breaks (optional, e.g. every 5 components)
  scale_x_continuous(breaks = seq(0, 90, by = 5)) +

  # Primary y-axis is variance; secondary y-axis is the cumulative proportion
  scale_y_continuous(
    name = "Variance (Eigenvalue)",
    sec.axis = sec_axis(
      trans = ~ . / max(pca_table_90$eigenvalue),
      name = "Cumulative Proportion of Variance Explained"
    )
  ) +

  # Labels and title
  labs(
    title = "Scree Plot For Principal Components",
    x = "Principal Component"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic"),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )

ggsave("../plots/scree_plot.png", width = 12, height = 8)
```

## 5.2 ROC Curves

```{r}
ggroc(list(
  benchmark = roc_benchmark,
  scree = roc_scree,
  greaterthan1 = roc_greaterthan1
)) +
  labs(
    title = "ROC Curves of Different Models",
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)",
    color = "Model"
  ) +
  geom_abline(
    intercept = 1,
    slope = 1,
    linetype = "dashed",
    color = "gray",
    size = 0.8
  ) +
  scale_color_manual(
    values = c("steelblue", "orange", "darkgreen"),
    labels = c("Benchmark", "Scree", "Greater Than 1")
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic"),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )

ggsave("../plots/roc_curves.png", width = 12, height = 8)
```

# 6. Iteratively Add Components

## 6.1 Estimate Logistic Regression Models

```{r}
logit_results <- rep(NA, length(pc_greaterthan1)) # to store results
logit_warnings <- rep(FALSE, length(pc_greaterthan1)) # to track warnings

for (i in seq_along(pc_greaterthan1)) {
  captured_warning <- NULL # set caputured_warning to NULL
  logit <- suppressWarnings(
    withCallingHandlers({
      # Fit logistic regression model
      logit <- data_pca %>%
        select(all_of(pc_greaterthan1[1:i]), class) %>%
        train(
          class ~ .,
          data = .,
          method = "glm",
          family = "binomial",
          trControl = trainControl(method = "LOOCV", savePredictions = "final", classProbs = TRUE)
        )
      # Store accuracy
      logit_results[i] <- logit$results$Accuracy
    }, warning = function(w) {
      captured_warning <<- w # Capture the warning
    })
  )

  # logit_warnings == T if warning occurred
  if (!is.null(captured_warning)) {
    logit_warnings[i] <- TRUE
    message(paste("Warning in iteration", i, ":", conditionMessage(captured_warning)))
  }
}
```

## 6.2 Plot Accuracy by Number of Components

```{r}
# Create a data frame to combine all data
plot_data <- data.frame(
  x = 1:length(pc_greaterthan1),
  y = logit_results,
  warning = ifelse(logit_warnings, "No", "Yes") # Create a factor for warnings
)

ggplot() +
  # Line for accuracy results
  geom_line(data = plot_data, aes(x = x, y = y)) +

  # Points with color mapped to 'warning'
  geom_point(
    data = plot_data,
    aes(x = x, y = y, color = warning), size = 3
  ) +

  labs(
    x = "Number of Components",
    y = "Accuracy",
    title = "Accuracy by Number of Principal Components",
    color = "Convergence" # Label for the legend
  ) +
  scale_color_manual(values = c("No" = "darkblue", "Yes" = "darkorange")) + # Custom colors
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic"),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold")
  )

ggsave("../plots/accuracy_by_components.png", width = 12, height = 8)
```
